2025 US School Shooting Predictions and Solutions

### 2024-11-28 paper/paper.qmd
update model section

library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)

#### Read data ####
data <- read_csv("../data/02-analysis_data/analysis_data.csv")
```

@@

- tidyverse [@tidyverse2024]
- dplyr [@dplyr2024]
- lubridate [@lubridate2024]
- readr [@readr2024]
- stringr [@stringr2024]
- arrow [@arrow2024]
- testthat [@testthat2024]
- randomForest [@randomForest2024]
- ggplot2
- maps

@@

$ Causalities = Killed + Injured $

@@

```{r}
#| label: fig-Number_of_Incidents_Over_Time
#| fig-cap: "Number of US School Shooting Incidents Over Time"
#| echo: false

@@

# Ensure state names are in lowercase to match map_data
state_counts <- state_counts %>%
  mutate(state = tolower(state))

@@

# Replace NA values in `total_casualties` with 0 (for states with no recorded casualties)
map_data$total_casualties[is.na(map_data$total_casualties)] <- 0

ggplot(map_data, aes(long, lat, group = group, fill = total_casualties)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(
    low = "#E6F7FF",
    high = "#0072B2",
    name = "Total Casualties",
    breaks = c(0, 50, 100, 150, 200)  # Adjust the breaks based on your data
  ) +
  labs(
    title = "US School Shooting Casualties by State",
    subtitle = "Number of casualties visualized geographically",
    x = NULL,
    y = NULL
  ) +
  coord_fixed(1.3) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold", size = 13, margin = ggplot2::margin(t = 10, b = 10)),
    plot.subtitle = ggplot2::element_text(hjust = 0.5, size = 10, margin = ggplot2::margin(t = 0, b = 20)),
    legend.position = "right",  # Alternative: Use "bottom" for a horizontal legend
    legend.key.height = unit(1, "cm"),  # Adjust legend size
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    plot.margin = ggplot2::margin(t = 20, r = 20, b = 20, l = 20)  # Increase margins for PDF export
  )

@@

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

>>>>

library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)
library(knitr)
library(randomForest)
library(here)
#### Read data ####
data <- read_csv("../data/02-analysis_data/analysis_data.csv")
```

@@

- tidyverse [@tidyverse2024]
- dplyr [@dplyr2024]
- lubridate [@lubridate2024]
- readr [@readr2024]
- stringr [@stringr2024]
- arrow [@arrow2024]
- testthat [@testthat2024]
- randomForest [@randomForest2024]
- ggplot2
- maps
- knitr

@@

In this dataset, there are three variables that have relationship: $ Causalities = Killed + Injured $.

@tbl-casualties_summary_statistics summarizes the mean, median and standard deviation of casualties from school shooting events in US from 1999 to today. The standard deviation of 3.72 indicates that while most incidents have casualties close to the mean, there is a wide range of variability, with some incidents having significantly higher number of casualties.

```{r}
#| label: tbl-casualties_summary_statistics
#| fig-cap: "Summary Statistics for Casualties"
#| echo: false

# Ensure the data contains the 'casualties' column
# Calculate the mean, median, and standard deviation
casualties_summary <- data %>%
  summarise(
    Mean = mean(casualties, na.rm = TRUE),
    Median = median(casualties, na.rm = TRUE),
    `Standard Deviation` = sd(casualties, na.rm = TRUE)
  )

# Display the summary in a beautiful table
kable(casualties_summary, caption = "Summary Statistics for Casualties") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

@@

@@

# Ensure state names in state_counts are lowercase to match map_data
state_counts <- state_counts %>%
  mutate(state = tolower(state))

@@

# Replace missing values in `total_casualties` with 0 for states with no data
map_data$total_casualties[is.na(map_data$total_casualties)] <- 0

# Plot the map
ggplot(map_data, aes(long, lat, group = group, fill = total_casualties)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(
    low = "#E6F7FF",  # Light blue
    high = "#0072B2",  # Dark blue
    name = "Total Casualties",
    breaks = c(0, 50, 100, 150, 200)  # Adjust breaks based on your data
  ) +
  labs(
    title = "US School Shooting Casualties by State",
    subtitle = "Number of casualties visualized geographically",
    x = NULL,
    y = NULL
  ) +
  coord_fixed(1.3) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold", size = 13, margin = ggplot2::margin(t = 10, b = 10)),
    plot.subtitle = ggplot2::element_text(hjust = 0.5, size = 10, margin = ggplot2::margin(t = 0, b = 20)),
    legend.position = "right",  # Alternative: Use "bottom" for a horizontal legend
    legend.key.height = unit(1, "cm"),  # Adjust legend size
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    plot.margin = ggplot2::margin(t = 20, r = 20, b = 20, l = 20)  # Increase margins for PDF export
  )

@@

Predictor variables in this analysis is state and year.

 - State represents the state where the schools are located.

 - Year represents the shooting year.

```{r}
#| label: fig-Number_of_Incidents_Over_Time
#| fig-cap: "Number of US School Shooting Incidents Over Time"
#| echo: false
#| eval: false

@@

## Model Set-Up

To model the total number of casualties resulting from school shooting incidents, a **Random Forest regression model** is employed. The model is designed to predict casualties based on state-level and temporal features derived from the dataset. The Random Forest algorithm is chosen for its robustness to non-linear relationships, its ability to handle interactions among predictors, and its ability to integrate categorical variables without creating dummy variables.

The mathematical representation of the model is: $C = f(X_1, X_2, ..., X_p) + \epsilon$

where:

- $C$: The total number of casualties (sum of fatalities and injuries) for a given year and state.  
- $f(\cdot)$: The function estimated by the Random Forest model, representing the aggregate prediction from all decision trees.  
- $X_1, X_2, ..., X_p$: Predictors, including:  
  - $X_1$: Year (numeric), capturing temporal trends.  
  - $X_2$: State (categorical), capturing regional effects.  
- $\epsilon$: The error term, representing unobserved factors affecting the number of casualties.

## Model Justification
- **Year** is included to account for temporal trends in school shootings, reflecting potential increases or decreases over time.
- **State** is treated as a categorical variable to capture location-specific effects, such as differences in legislation, policing, or socioeconomic factors.

The choice of these features aligns with the data section, ensuring that variables with potential predictive value are incorporated without overfitting or introducing unnecessary complexity.

### Assumptions and Limitations
- **Assumptions**:
  1. Casualties are conditionally independent given the predictors.
  2. The relationship between predictors and casualties can be approximated by the ensemble of decision trees.
  3. Data is representative of underlying trends and free of major sampling biases.

- **Limitations**:
  1. The model may not generalize well to future data if underlying trends shift dramatically (e.g., policy changes or societal events).
  2. Random Forests lack explicit interpretability compared to simpler models.

### Software Implementation
The model is implemented in **R**, using the randomForest [@randomforest] package for training and evaluation.

### Model Validation
- **Train/Test Splits**:
  The data is split into training (80%) and test (20%) sets. The model is trained on the former and validated on the latter to ensure generalizability.

- **Evaluation Metrics**:
  - **Root Mean Squared Error (RMSE)** and **Mean Absolute Error (MAE)** are calculated to assess prediction accuracy.
  - Final metrics:
    - RMSE: 1.491141 casualties
    - MAE: 1.096164 casualties

### Alternative Models Considered
- **Linear Regression**:
  - Strengths: Simplicity and interpretability.
  - Weaknesses: Inadequate for capturing non-linear relationships and interactions present in the data.

- **Gradient Boosting Machines (GBM)**:
  - Strengths: Often more accurate due to its iterative learning.
  - Weaknesses: Higher risk of overfitting, more computationally expensive and less intepretibility.

The Random Forest model was chosen over these alternatives for its balance of flexibility, robustness, and interpretability of feature importance.

@@

```{r}
#| echo: false
#| eval: true
#| label: fig-predicted_shooting_visualization
#| tbl-cap: "Predicted 2025 US School Shootings Numbers"
#| warning: false

#### Step 1: Load the Data and Model ####
# Load cleaned data
clean_data <- read.csv("../data/02-analysis_data/no_outlier_data.csv")

# Ensure 'state' is a factor
clean_data$state <- as.factor(clean_data$state)

# Load the Random Forest model
rf_model_clean <- readRDS(file = here::here("models/rf_model.rds"))

#### Step 2: Prepare Data for Predictions ####
# Year(s) for prediction
years <- 2025

# Create a data frame with unique states and years
prediction_data <- expand.grid(
  year = years,
  state = unique(clean_data$state)  # Use unique states
)

# Convert 'state' to a factor and match levels with training data
prediction_data$state <- factor(prediction_data$state, levels = levels(clean_data$state))

# Ensure no missing levels
if (any(is.na(prediction_data$state))) {
  stop("State levels in prediction_data do not match clean_data.")
}

# Make predictions
predictions <- predict(rf_model_clean, newdata = prediction_data)

# Add predictions to the data frame
prediction_data$predicted_casualties <- predictions

#### Step 3: Prepare Map Data ####
# Get US map data
us_map <- map_data("state")

# Convert state names in `us_map` to lowercase
us_map$region <- tolower(us_map$region)

# Convert state in `prediction_data` to lowercase for compatibility
prediction_data$state <- tolower(as.character(prediction_data$state))

# Merge map data with predictions
map_data <- us_map %>%
  left_join(prediction_data, by = c("region" = "state"))

# Fill missing predictions with 0 (optional, avoids grey states)
map_data$predicted_casualties[is.na(map_data$predicted_casualties)] <- 0

#### Step 4: Plot the Map ####
ggplot(map_data, aes(long, lat, group = group, fill = predicted_casualties)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(
    low = "#E6F7FF",  # Light blue
    high = "#0072B2", # Dark blue
    name = "Casualties"
  ) +
  labs(
    title = "Predicted US School Shooting Casualties by State (2025)",
    subtitle = "Number of casualties visualized geographically",
    x = NULL,
    y = NULL
  ) +
  coord_fixed(1.3) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    legend.position = "right",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
